---
title: EM-DAT R Tutorial 2
output: html_document
---

# Making Maps

If you have followed the [EM-DAT R Tutorial 1](./r_tutorial_1_basic_operations_and_plotting.ipynb) or are already familiar with [`dplyr`](https://dplyr.tidyverse.org/) and [`ggplot2`](https://ggplot2.tidyverse.org/), this second tutorial will show you basic examples on how to make maps with the EM-DAT data using the [`sf`](https://r-spatial.github.io/sf/) R package.

**Note**: This tutorial is also available on the [EM-DAT Documentation Website](https://doc.emdat.be/docs/additional-resources-and-tutorials/).

## Load Packages

Let us load the necessary packages and print their versions. For this tutorial, we used `sf` v.1.1-2, `dplyr` v.1.2.0, `jsonlite` v.2.0.0, and `ggplot2` v.4.0.2. If your package versions are different, you may have to adapt this tutorial by checking the corresponding package documentation.

```{r message=FALSE}
library(readxl)    # read Excel files
library(dplyr)     # data manipulation
library(tidyr)     # data reshaping
library(sf)        # simple features: spatial vector data
library(jsonlite)  # parse JSON strings
library(ggplot2)   # plotting library
set_theme(theme_minimal())
```

## Creating a World Map

To create a world map, we need the EM-DAT data and a shapefile containing the country geometries.

**EM-DAT**: We download and load the EM-DAT data using `readxl`. As in Tutorial 1, we use the [EM-DAT Archive published on Dataverse](https://doi.org/10.14428/DVN/I0LTPH).

**Country Shapefile**: We download a country shapefile from [Natural Earth Data](https://www.naturalearthdata.com/). For a world map, we download the low resolution [1:110m Admin 0 - Countries](https://www.naturalearthdata.com/downloads/110m-cultural-vectors/) and unzip it.

### Load and Filter EM-DAT

Let us load EM-DAT and filter it to make a global map of Earthquake disasters between 2000 and 2023. We calculate the number of unique identifiers (`DisNo.`) per country (`ISO`). We refer to the standard `ISO` column instead of the `Country` column of the [EM-DAT Public Table](https://doc.emdat.be/docs/data-structure-and-content/emdat-public-table/) to be able to make a join with the country shapefile.

```{r}
# modify file name or path as needed
df <- read_excel("data/260430_emdat_archive.xlsx", guess_max = Inf)

earthquake_counts <- df |>
  filter(
    `Disaster Type` == "Earthquake",
    `Start Year` >= 2000,
    `Start Year` < 2024
  ) |>
  count(ISO, name = "EarthquakeCount")

earthquake_counts
```

### Load the Country Shapefile

We use the [`st_read`](https://r-spatial.github.io/sf/reference/st_read.html) function to load the country shapefile and parse it into an `sf` object. An `sf` object is similar to a `data.frame`, except that it has a geometry column.

We provide the `dsn` argument, which is either a file name if located in the same directory as the running script, or a relative or absolute path, if not. In our case the shapefile with the `.shp` extension is located in the `ne_110m_admin_0_countries` folder.

Since the object contains 168 attribute columns, we only keep the one we are interested in, i.e., `ISO_A3`. The `geometry` column is *sticky*: `sf` keeps it automatically through a `select`, so it does not need to be named.

**Note**: We display `sf` objects with [`print`](https://r-spatial.github.io/sf/reference/print.html) throughout this tutorial. This gives the standard `sf` summary, reporting the number of features, the geometry type, the bounding box, and the coordinate reference system, all of which are worth checking after loading spatial data.

```{r}
gdf <- st_read(
  "data/ne_110m_admin_0_countries/ne_110m_admin_0_countries.shp",  # <-- Change path if necessary
  quiet = TRUE
)

gdf <- gdf |> select(ISO_A3)
print(gdf)
```

**Important Notice**: Above, some geometries do not have a valid ISO code and are coded `-99` instead. Below, you will see that some ISO in EM-DAT are not matched with a geometry. Beyond this basic tutorial, we advise to carefully evaluate these correspondences and non-correspondences between ISO codes and to read the [EM-DAT Documentation about ISO codes](https://doc.emdat.be/docs/data-structure-and-content/spatial-information/).

### Join the Two Datasets

Let us merge the two datasets with a full join, using the [`full_join`](https://dplyr.tidyverse.org/reference/mutate-joins.html) function. We prefer a full join to keep the geometries of countries for which EM-DAT has no records.

**Note**: The join keys have different names in the two tables, so we pair them with [`join_by`](https://dplyr.tidyverse.org/reference/join_by.html). Starting the pipe from `gdf` matters: the result keeps the `sf` class, and stays plottable as a map.

```{r}
earthquake_counts_with_geom <- gdf |>
  full_join(earthquake_counts, by = join_by(ISO_A3 == ISO))

print(earthquake_counts_with_geom)
```

### Make the Map

To make the map, we use `ggplot2` together with the [`geom_sf`](https://ggplot2.tidyverse.org/reference/ggsf.html) layer, which understands the geometry column of an `sf` object and picks a sensible map projection automatically.

We map the `EarthquakeCount` column to the `fill` aesthetic, and use [`scale_fill_distiller`](https://ggplot2.tidyverse.org/reference/scale_brewer.html) to apply a sequential `Reds` colour palette. Countries without records are drawn in light grey via the `na.value` argument.

```{r}
ggplot(earthquake_counts_with_geom) +
  geom_sf(aes(fill = EarthquakeCount), colour = "white", linewidth = 0.1) +
  scale_fill_distiller(
    palette = "Reds",
    direction = 1,
    na.value = "lightgrey",
    name = "Nb of EM-DAT\nEarthquake"
  ) +
  labs(x = "Lon.", y = "Lat.")
```

## Creating a Map at Admin Level 1

We can create a more detailed map using the `GADM Admin Units` column in the [EM-DAT Public Table](https://doc.emdat.be/docs/data-structure-and-content/emdat-public-table/). This column contains the identifiers of administrative units of level 1 or 2 as defined by the [Database of Global Administrative Areas (GADM), version 4.1](https://gadm.org/), for countries impacted by non-biological natural hazards.

**Note**: EM-DAT also carries a legacy `Admin Units` column referring to the Global Administrative Unit Layer (GAUL). GAUL is no longer supported, and new work should use the `GADM Admin Units` column shown here.

Similarly to the country map, we need a file containing the GADM geometries. GADM publishes one geopackage per country, so we only download [the Japanese one](https://geodata.ucdavis.edu/gadm/gadm4.1/gpkg/gadm41_JPN.gpkg) (~33 MB), since this tutorial focuses on Japanese earthquake occurrence in EM-DAT.

### Load the GADM Geopackage

The file is a geopackage `.gpkg` that contains multiple layers. Let us first describe these layers with the [`st_layers`](https://r-spatial.github.io/sf/reference/st_layers.html) function.

```{r}
print(st_layers("data/gadm41_JPN.gpkg"))
```

- The `ADM_ADM_0` layer contains the country geometry.
- The `ADM_ADM_1` layer contains the 47 Japanese prefectures. This is the level we map here.
- The `ADM_ADM_2` layer contains the 1811 municipalities.

GADM ships a ready-made administrative level 1 layer, so we can load it directly. Let us keep only the identifier `GID_1` and the name `NAME_1`.

```{r}
gadm_adm1 <- st_read("data/gadm41_JPN.gpkg", layer = "ADM_ADM_1", quiet = TRUE)

gadm_adm1 <- gadm_adm1 |> select(GID_1, NAME_1)
print(head(gadm_adm1))
```

We will also generate a dictionary (as a vector of named values) to retrieve the `GID_1` from the `GID_2`, at a later stage.

```{r}
gadm_adm2 <- st_read("data/gadm41_JPN.gpkg", layer = "ADM_ADM_2", quiet = TRUE)

if (any(duplicated(gadm_adm2$GID_2))) {
  warning("Duplicated keys found in GID_2 identifiers")
}

# applying the GID_2 value as name for each corresponding value of GID_1
gid_2to1 <- with(gadm_adm2, setNames(GID_1,GID_2))
```

### Filter EM-DAT Data

```{r}
df_jpn <- df |>
  filter(
    ISO == "JPN",
    `Disaster Type` == "Earthquake",
    `Start Year` >= 2000,
    `Start Year` < 2024
  ) |>
  select(`DisNo.`, `GADM Admin Units`)

df_jpn
```

**Note**: Some events are not geolocated at a finer administrative level and have a missing (`NA`) value.

### Convert Admin 2 units to Admin 1 units

We create an R function, `gadm_to_admin1`, to extract the administrative level 1 identifiers from the `GADM Admin Units` column.

The `GADM Admin Units` column holds a JSON array whose entries provide either a level 1 identifier (`gid_1`) or a level 2 identifier (`gid_2`). GADM identifiers are *hierarchical strings*: the level 2 identifier `JPN.19.68_2` is nested under the level 1 identifier `JPN.19_1`. We could therefore derive the parent unit by trimming the string, with no spatial lookup required.

Yet, it is still preferable to avoid string manipulation and rely on relational matches in automated settings, using the dictionary previously generated.

```{r}
#' Convert a GADM Admin Units JSON string to administrative level 1 identifiers.
#'
#' @param json_str A JSON string representing administrative areas, or NA.
#' @return A character vector of unique GADM level 1 (GID_1) identifiers.
gadm_to_admin1 <- function(json_str) {
  if (is.na(json_str)) {
    return(character(0))
  }
  entries <- fromJSON(json_str, simplifyVector = FALSE)
  ids <- vapply(entries, function(entry) {
    if (!is.null(entry$gid_1)) {
      return(entry$gid_1)
    } else if (!is.null(entry$gid_2)) {
      # dictionary lookup from the GID_2
      gid_1 <- gid_2to1[entry$gid_2]
      if (!is.na(gid_1)) {
        return(gid_1)
      }
    }
    # fail if nothing was returned
    stop(
      "Administrative code is missing from the provided data or could not be matched against referential."
    )
  }, character(1))
  return(unique(ids))
}
```

We apply the function to all elements of the `GADM Admin Units` column. Because each event may span several administrative units, the function returns a vector of varying length, which we store in a *list column* using [`lapply`](https://rdrr.io/r/base/lapply.html).

**Note**: A list column holds a whole vector in each cell, so it has no fixed width and cannot be shown as an ordinary table. Below we collapse each list into a single comma-separated string purely for display; `df_jpn` itself keeps the list column, which is what the next step needs.

```{r}
df_jpn <- df_jpn |>
  mutate(GID_1 = lapply(`GADM Admin Units`, gadm_to_admin1))

# Collapse the list column to a string, for display only
df_jpn |>
  select(`DisNo.`, GID_1) |>
  mutate(GID_1 = sapply(GID_1, paste, collapse = ", "))
```

### Count Earthquakes per Admin 1 Units

This can be done by applying the [`unnest_longer`](https://tidyr.tidyverse.org/reference/unnest_longer.html) function on the new `GID_1` list column. The function adds one row per identifier contained in each list. Then the counting can be performed using the former `count` approach.

**Note**: Events with no administrative unit hold an empty vector and are dropped by `unnest_longer`.

```{r}
count_per_adm1 <- df_jpn |>
  unnest_longer(GID_1) |>
  count(GID_1, name = "EQ Count")

head(count_per_adm1)
```

### Join the Two Datasets

Again, we can use a join to bring the two datasets together, here on the shared `GID_1` column. We use a [`left_join`](https://dplyr.tidyverse.org/reference/mutate-joins.html) starting from the geometries, so that every prefecture is kept on the map, including those never affected.

```{r}
gdf_jpn_adm1_merged <- gadm_adm1 |>
  left_join(count_per_adm1, by = "GID_1")

print(gdf_jpn_adm1_merged)
```

### Make the Map

```{r}
ggplot(gdf_jpn_adm1_merged) +
  geom_sf(aes(fill = `EQ Count`), linewidth = 0.1) +
  scale_fill_distiller(
    palette = "YlOrRd",
    direction = 1,
    na.value = "white",
    name = "Earthquake Count\nin EM-DAT"
  ) +
  labs(x = "Longitude", y = "Latitude")
```

We have just covered the basics on how to join the EM-DAT data with an [`sf`](https://r-spatial.github.io/sf/) spatial object to make maps. To delve further into your analyses, we encourage you to continue your learning of [`sf`](https://r-spatial.github.io/sf/) and [`ggplot2`](https://ggplot2.tidyverse.org/), or, in particular, [`tmap`](https://r-tmap.github.io/tmap/) for more advanced map customization, with the many resources available online. The book [Geocomputation with R](https://r.geocompx.org/) is a good place to start.

